---
title: "Adding metadata"
author: "Thierry Onkelinx"
output:
  rmarkdown::html_vignette:
        fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Adding metadata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

`git2rdata` supports extra metadata since version 0.4.1.
Metadata is stored in a separate file with the same name as the data file, but with the extension `.yml`.
The metadata file is a YAML file with a specific structure.
The metadata file contains a generic section and a section for each field in the data file.
The generic section contains information about the data file as a whole.
The fields sections contain information about the fields in the data file.
The metadata file is stored in the same directory as the data file.

The generic section contains the following mandatory properties, automatically created by `git2rdata`:

- `git2rdata`: the version of `git2rdata` used to create the metadata.
- `datahash`: the hash of the data file.
- `hash`: the hash of the metadata file.
- `optimize`: a logical indicating whether the data file is optimized for `git2rdata`.
- `sorting`: a character vector with the names of the fields in the data file.
- `split_by`: a character vector with the names of the fields used to split the data file.
- `NA string`: the string used to represent missing values in the data file.

The generic section can contain the following optional properties:

- `table name`: the name of the dataset.
- `title`: the title of the dataset.
- `description`: a description of the dataset.

The fields sections contain the following mandatory properties, automatically created by `git2rdata`:

- `type`: the type of the field.
- `class`: the class of the field.
- `levels`: the levels of the field (for factors).
- `index`: the index of the field (for factors).
- `NA string`: the string used to represent missing values in the field.

The fields sections can contain the following optional properties:

- `description`: a description of the field.

## Adding metadata

`write_vc()` only stores the mandatory properties in the metadata file.

```{r store-metadata}
library(git2rdata)
root <- tempfile("git2rdata-metadata")
dir.create(root)
write_vc(iris, file = "iris", root = root, sorting = "Sepal.Length")
```

## Reading metadata

`read_vc()` reads the metadata file and adds it as attributes to the `data.frame`.
`print()` and `summary()` alert the user to the `display_metadata()` function.
This function displays the metadata of a `git2rdata` object.
Missing optional metadata results in an `NA` value in the output of `display_metadata()`.

```{r read-metadata}
my_iris <- read_vc("iris", root = root)
str(my_iris)
print(head(my_iris))
summary(my_iris)
display_metadata(my_iris)
```

## Updating the optional metadata

To add metadata to a `git2rdata` object, use the `update_metadata()` function.
This function allows you to add or update the optional metadata of a `git2rdata` object.
Setting an argument to `NA` or an empty string will remove the corresponding property from the metadata.
The function only updates the metadata file, not the data file.
To see the changes, read the object again before using `display_metadata()`.
Note that all the metadata is available in the `data.frame` as attributes.

```{r update-metadata}
update_metadata(
  file = "iris", root = root, name = "iris", title = "Iris dataset",
  description =
"The Iris dataset is a multivariate dataset introduced by the British
statistician and biologist Ronald Fisher in his 1936 paper The use of multiple
measurements in taxonomic problems.",
  field_description = c(
    Sepal.Length = "The length of the sepal in cm",
    Sepal.Width = "The width of the sepal in cm",
    Petal.Length = "The length of the petal in cm",
    Petal.Width = "The width of the petal in cm",
    Species = "The species of the iris"
  )
)
my_iris <- read_vc("iris", root = root)
display_metadata(my_iris)
str(my_iris)
```

