---
title: "Getting Started Storing Dataframes as Plain Text"
author: "Thierry Onkelinx"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started Storing Dataframes as Plain Text}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  comment = "#>"
)
options(width = 83)
```

## Introduction

This vignette motivates why we wrote `git2rdata` and illustrates how you can use it to store dataframes as plain text files.

### Maintaining Variable Classes

R has different options to store dataframes as plain text files from R. 
Base R has `write.table()` and its companions like `write.csv()`. 
Some other options are `data.table::fwrite()`, `readr::write_delim()`, `readr::write_csv()` and `readr::write_tsv()`. 
Each of them writes a dataframe as a plain text file by converting all variables into characters. 
After reading the file, they revert this conversion. 
The distinction between `character` and `factor` gets lost in translation.
`read.table()` converts by default all strings to factors, `readr::read_csv()` keeps by default all strings as character. 
These functions cannot recover the factor levels.
These functions determine factor levels based on the observed levels in the plain text file. 
Hence factor levels without observations will disappear. 
The order of the factor levels is also determined by the available levels in the plain text file, which can be different from the original order.

The `write_vc()` and `read_vc()` functions from `git2rdata` keep track of the class of each variable and, in case of a factor, also of the factor levels and their order.
Hence this function pair preserves the information content of the dataframe. The `vc` suffix stands for
**v**ersion **c**ontrol<!-- spell-check: ignore -->
as these functions use their full capacity in combination with a version control system.

## Efficiency Relative to Storage and Time

### Optimizing File Storage

Plain text files require more disk space than binary files. 
This is the price we have to pay for a readable file format. 
The default option of `write_vc()` is to create file as compact as possible. 
Since we use a tab delimited file format, we can omit quotes around character variables. 
This saves 2 bytes per row for each character variable. 
`write_vc` add quotes automatically in the exceptional cases when we needed them, e.g. to store a string that contains tab or newline characters.
We don't add quotes to row-variable combinations where we don't need them.

Since we store the class of each variable, we can further reduce the file size by following rules:

- Store a `logical` as 0 (FALSE), 1 (TRUE) or NA to the data.
- Store a `factor` as its indices in the data. 
Store the index, labels of levels and their order in the metadata.
- Store a `POSIXct` as a numeric to the data. 
Store the class and the origin in the metadata. 
Store and return timestamps as UTC.
- Store a `Date` as an integer to the data. 
Store the class and the origin in the metadata.

Storing the factors,
POSIXct <!-- spell-check: ignore -->
and Date as their index, makes them less user readable. The user can turn off this optimization when user readability is more important than file size.

### Optimized for Version Control

Another main goal of `git2rdata` is to optimise the storage of the plain text files under version control. `write_vc()` and `read_vc()` has methods for interacting with [git](https://git-scm.com/) repositories using the `git2r` framework. Users who want to use git without `git2r` or use a different version control system (e.g. [Subversion](https://subversion.apache.org/), [Mercurial](https://www.mercurial-scm.org/)), still can use `git2rdata` to write the files to disk and uses their preferred workflow on version control. 

Hence, `write_vc()` will always perform checks to look for changes which potentially lead to large diffs. More details on this in `vignette("version_control", package = "git2rdata")`. Some problems will always yield a warning. Other problems will yield an error by default. The user can turn these errors into warnings by setting the `strict = FALSE` argument.

As this vignette ignores the part on version control, we will always use `write_vc(strict = FALSE)` and hide the warnings to improve the readability.

## Basic Usage

Let's start by setting up the environment. We need a directory to store the data and a dataframe to store.

```{r}
# Create a directory in tempdir
path <- tempfile(pattern = "git2r-")
dir.create(path)
# Create dummy data
set.seed(20190222)
x <- data.frame(
  x = sample(LETTERS),
  y = factor(
    sample(c("a", "b", NA), 26, replace = TRUE),
    levels = c("a", "b", "c")
  ),
  z = c(NA, 1:25),
  abc = c(rnorm(25), NA),
  def = sample(c(TRUE, FALSE, NA), 26, replace = TRUE),
  timestamp = seq(
    as.POSIXct("2018-01-01"),
    as.POSIXct("2019-01-01"),
    length = 26
  ),
  stringsAsFactors = FALSE
)
str(x)
```

## Storing Optimized

Use `write_vc()` to store the dataframe. 
The `root` argument refers to the base directory where we store the data. 
The `file` argument becomes the base name of the files. 
The data file gets a `.tsv` extension, the metadata file a `.yml` extension. 
`file` can include a relative path starting from `root`.

```{r first_write}
library(git2rdata)
write_vc(x = x, file = "first_test", root = path, strict = FALSE)
```

`write_vc()` returns a vector of relative paths to the raw data and metadata files. 
The names of this vector contains the hashes of these files. 
We can have a look at both files. 
We'll display the first 10 rows of the raw data. 
Notice that the YAML format of the metadata has the benefit of being both human and machine readable. 

```{r manual_data}
print_file <- function(file, root, n = -1) {
  fn <- file.path(root, file)
  data <- readLines(fn, n = n)
  cat(data, sep = "\n")
}
print_file("first_test.tsv", path, 10)
print_file("first_test.yml", path)
```


## Storing Verbose

Adding `optimize = FALSE` to `write_vc()` will keep the raw data in a human readable format.
The metadata file is slightly different.
The most obvious is the `optimize: no` tag and the different hash.
Another difference is the metadata for
POSIXct <!-- spell-check: ignore -->
and Date classes.
They will no longer have an origin tag but a format tag.

Another important difference is that we store the data file as comma separated values instead of tab separated values.
We noticed that the `csv` file format is more easily recognised by a larger audience as a data file.


```{r write_verbose}
write_vc(x = x, file = "verbose", root = path, optimize = FALSE, strict = FALSE)
```

```{r manual_verbose_data}
print_file("verbose.csv", path, 10)
print_file("verbose.yml", path)
```

## Efficiency Relative to File Storage

Storing dataframes optimized or verbose has an impact on the required file size. 
The [efficiency](efficiency.html#on-a-file-system) vignette give a comparison.

## Reading Data

You retrieve the data with `read_vc()`. 
This function will reinstate the variables to their original state.

```{r first_read}
y <- read_vc(file = "first_test", root = path)
all.equal(x, y, check.attributes = FALSE)
y2 <- read_vc(file = "verbose", root = path)
all.equal(x, y2, check.attributes = FALSE)
```

`read_vc()` requires the meta data. 
It cannot handle dataframe not stored by `write_vc()`.

## Missing Values

`write_vc()` has an `na` argument which specifies the string which to use for missing values. 
Because we avoid using quotes, this string must be different from any character value in the data. 
This includes factor labels with verbose data storage. 
`write_vc()` checks this and will always return an error, even with `strict = FALSE`.

```{r echo = FALSE, results = "hide"}
stopifnot("X" %in% x$x, "b" %in% x$y)
```

```{r na_string, error = TRUE}
write_vc(x, "custom_na", path, strict = FALSE, na = "X", optimize = FALSE)
write_vc(x, "custom_na", path, strict = FALSE, na = "b", optimize = FALSE)
write_vc(x, "custom_na", path, strict = FALSE, na = "X")
write_vc(x, "custom_na", path, strict = FALSE, na = "b")
```

Please note that `write_vc()` uses the same  NA string for the entire dataset, thus for every variable. 

```{r manual_na_data}
print_file("custom_na.tsv", path, 10)
print_file("custom_na.yml", path, 4)
```

The default string for missing values is `"NA"`. We recommend to keep this default, as long as the dataset permits it. A first good alternative is an empty string (`""`). If that won't work either, you'll have to use your imagination. Try to keep it short, clear and robust^[robust in the sense that you won't need to change it later].

```{r empty_na}
write_vc(x, "custom_na", path, strict = FALSE, na = "")
```
